Future challenges for Research Software Engineering

### 2025-05-16 future_challenges_for_RSEng.tex
typo

\subsection{Definition and Evolution of RSEs}
\begin{whatis}{}{definition-evolution-rse}
The term RSE is established in the minds of university boards and leadership, as a valuable member in the academic framework.
New RSEs are trained through a specialized master's program, and their support is valued by researchers. We still teach researchers basic software skills, but the content of this lecture has changed drastically.
\end{whatis}
% Here we consider how the topics of RSEng will change in the face of the digitalisation of society.
% Also here: What will our topics be after version control doesn't need to be taught anymore?
% Digitalisation will also feature more heavily in the domain curricula.
In this discussions the workshop participants found that
the digital skills of domain scientists will increase,
and software will become easier to interact with,
reducing reliance on RSEs to train newcomers on the most basic tasks.
This means, that today's workshop content of proper versioning and data management could become obsolete
by more intuitive tooling, hence basic courses can focus on more higher-level concepts.
User training sessions might focus on more advanced topics
that cater to a smaller audience, or involve mentoring more frequently.

@@

\subsection{Development of the RSE community}
\begin{whatis}{}{development-rse-community}
National RSE communities are increasingly collaborating with local partners, fostering stronger networks across countries. Regular international conferences further amplify these connections, promoting the global exchange of ideas. The introduction of RSE Master's programs, coupled with a shift in culture towards valuing Research Software, has significantly elevated its profile. This growing recognition has led to a surge in membership within national societies, reflecting the increasing importance of RSEng in the research landscape.
\end{whatis}
In recent years, we have seen the formation of RSE institutes and advocacy groups, such as
the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
the US Research Software Sustainability Institute\footnote{\url{https://urssi.us}},
the European Virtual Institute for Research Software Excellence\footnote{\url{https://everse.software}},
and the EOSC Opportunity Area Expert Group 7 on Research Software%
\footnote{\url{https://eosc.eu/opportunity-area-exp/oa7-research-software}}.
We expect these initiatives to help increase the visibility and accelerate
the professionalisation%
of the RSE role.
In addition, a multitude of RSE communities self-organised at the national
and regional level to support more viable RSE career paths in academia.
We can expect RSEs to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
The community of RSE's will get a significant boost in their acceptance if RSE Masters programmes
become available at various universities.
With increased numbers, the visibility of the RSE community to industry will be enhanced.
\begin{story}{Development of RSE community}{RSEcommunity}
Kay stood proudly at the podium of the first International Research Software Engineering Conference held in the Southern Hemisphere, a historic moment for both her career and the global RSE community. Having moved to South Africa, she spoke passionately about the unique challenges researchers in the Global South face—limited access to high-performance computing, unreliable funding, and the persistent digital divide. Her keynote highlighted how local innovation thrives despite constraints, often out of necessity rather than choice. The audience, a diverse mix of global experts, listened intently as she called for equitable collaborations, not just inclusion as an afterthought.
\end{story}

@@

% society implications
The work being undertaken by RSEs may well have direct, or indirect,
ethical implications in areas such as social engineering \cite{s2erc,siadati2024},
climate \cite{Lannelongue2023} and more.
Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
Like engineers, RSEs take the responsibility for the software they develop.
RSEs need to be aware of these issues and develop their own ``red lines''.

@@

A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
This can then form the basis for institutional policies and guidelines as well as serve as a background for individual developers to consider how software can be used.

@@

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{}{academic}
We expect research software to be a central part of scientific practice on the same level as data and publications.
This will be supported by the first generations of RSE-Master graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, that are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
of academic staff at any level.
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects \cite{Kempf2025-draft}.
We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
At universities which have them, we want to see that researchers naturally accept them as part of their available services.
Additionally researchers will seamlessly interact with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative \cite{wzvgbayern}, has organisationally often adopted time-scales for change that are more reminiscent to institutions thinking on the time scale of generations.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref} \cite{hiddenref} hints at the potential for specialisation of research tasks. This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
We have already seen some specialisation trends in recent years such as
the HPC-RSE and AI-RSE which are quite recent developments, whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
But we could imagine new RSE specialisations emerging in the fields
of quantum computing, such as the Quantum-RSE, and in federated online platforms for research.
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

On the other hand the existence of people with a certain specilised skillset opens up possiblities for pooling.
Universities will benefit from pooling these specialists into RSE units, and offer their services to researchers.
Hopefully, these RSE-units are endowed with enough trust by the university boards that translates into stable funding.
With longer-term funding and more trust to pick up longer term projects,
some RSEs can focus on code modernisation, cleanup, and modularisation,
which can pay off in faster and easier development in the future.
This is an opportunity that would not be possible if relying on short-term funded solo RSEs.

>>>>

\subsection{Definition and Evolution of RSEs}
\begin{whatis}{}{definition-evolution-rse}
The term RSE is established in the minds of university boards and leadership, as a valuable member in the academic framework.
New RSEs are trained through a specialized master's program, and their support is valued by researchers. We still teach researchers basic software skills, but the content of this lecture has changed drastically.
\end{whatis}
% Here we consider how the topics of RSEng will change in the face of the digitalisation of society.
% Also here: What will our topics be after version control doesn't need to be taught anymore?
% Digitalisation will also feature more heavily in the domain curricula.
In this discussions the workshop participants found that
the digital skills of domain scientists will increase,
and software will become easier to interact with,
reducing reliance on RSEs to train newcomers on the most basic tasks
(JN: for example, versioning and archiving?)
(GC: Higher-level and more intuitive workflows that require less training).
User training sessions might focus on more advanced topics
that cater to a smaller audience, or involve mentoring more frequently.

@@

\subsection{Development of the RSE community}
\begin{whatis}{}{development-rse-community}
National RSE communities are increasingly collaborating with local partners, fostering stronger networks across countries. Regular international conferences further amplify these connections, promoting the global exchange of ideas. The introduction of RSE Master's programs, coupled with a shift in culture towards valuing Research Software, has significantly elevated its profile. This growing recognition has led to a surge in membership within national societies, reflecting the increasing importance of RSEng in the research landscape.
\end{whatis}
In recent years, we have seen the formation of RSE institutes and advocacy groups, such as
the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
the US Research Software Sustainability Institute\footnote{\url{https://urssi.us}},
the European Virtual Institute for Research Software Excellence\footnote{\url{https://everse.software}},
and the EOSC Opportunity Area Expert Group 7 on Research Software%
\footnote{\url{https://eosc.eu/opportunity-area-exp/oa7-research-software}}.
We expect these initiatives to help increase the visibility and accelerate
the professionalisation%
\footnote{MM: do we want to decide for UK or US English? Or keep it mixed? Not being a native English speaker I don't know how it reads.}
of the RSE role.
In addition, a multitude of RSE communities self-organised at the national
and regional level to support more viable RSE career paths in academia.
We can expect RSEs to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
The community of RSE's will get a significant boost in their acceptance if RSE Masters programmes
become available at various universities.
With increased numbers, the visibility of the RSE community to industry will be enhanced.
\begin{story}{Development of RSE community}{RSEcommunity}
Kay stood proudly at the podium of the first International Research Software Engineering Conference held in the Southern Hemisphere, a historic moment for both her career and the global RSE community. Having moved to South Africa, she spoke passionately about the unique challenges researchers in the Global South face—limited access to high-performance computing, unreliable funding, and the persistent digital divide. Her keynote highlighted how local innovation thrives despite constraints, often out of necessity rather than choice. The audience, a diverse mix of global experts, listened intently as she called for equitable collaborations, not just inclusion as an afterthought.
\end{story}

@@

% society implications
The work being undertaken by RSEs may well have direct, or indirect, ethical implications in areas such as social engineering [TODO find source], climate [TODO find source] and more.
Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
Like engineers, RSEs take the responsibility for the software they develop.
RSEs need to be aware of these issues and develop their own ``red lines''.

@@

A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
This can then form the basis for institutional policies and guidelines.

MM-NOTE:
  - Importance of having personal “red lines” and awareness of how software can be used.

@@

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{}{academic}
We expect research software to be a central part of scientific practice on the same level as data and publications.
This will be supported by the first generations of RSE-Master graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, that are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
of academic staff at any level.
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects \cite{Kempf2025-draft}.
We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
At universities which have them, we want to see that researchers naturally accept them as part of their available services.
Additionally researchers will seamlessly interact with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative (cite WissZeitVG here), has organisationally often adopted time-scales for change that are more reminiscient to civil services like pension funds.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref} \cite{hiddenref} hints at the potential for specialisation of research tasks. This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
We have already seen some specialisation trends in recent years such as
the HPC-RSE and AI-RSE which are quite recent developments, whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
But we could imagine new RSE specialisations emerging in the fields
of quantum computing, such as the Quantum-RSE, and in federated online platforms for research.
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

On the other hand the existence of people with a certain specialised skillset opens up possiblities for pooling.
Universities will benefit from pooling these specialists into RSE units, and offer their services to researchers.
Hopefully, these RSE-units are endowed with enough trust by the university boards that translates into stable funding.
With longer-term funding and more trust to pick up longer term projects,
some RSEs can focus on code modernisation, cleanup, and modularisation,
which can pay off in faster and easier development in the future.
This is an opportunity that would not be possible if relying on short-term funded solo RSEs.
